An Open-Source Implementation of the CMPS Algorithm for Assessing  Similarity of Bullets

### 2021-08-19 CMPSpaper_writing/CMPSpaper_writing.Rmd
added contents for CMPS_diff, scaled CMPS score, and variance ratio; updated the implementation section to match the current implementation

2. **identify candidate positions:** For each basis segment a set of candidate registration  positions on the comparison signature is identified based on the segment's similarity to the reference signature. In a first step, the cross-correlation function of the segment to the reference is calculated, then a number of positions with high correlation values are identified as candidate positions. In case multiple segment lengths are considered, the length of each basis segment is doubled and these two steps are repeated. Only when candidate positions coincide (or are similar enough), they are considered further. \autoref{fig:segplots} and \autoref{fig:seg_all} illustrate these ideas.

@@

        -   If `npeaks.set` is an integer vector of length more than 1, for example `npeaks.set = c(5, 3, 1)`, the multi-segment lengths strategy will be used: find the top five peaks at the first level, double the segment length and recalculate the cross-correlation function, then identify 3 peaks at the second level and repeat the process for a basis segment that is doubled again to identify 1 peak at the third level. \autoref{fig:seg_all} shows an example of three levels of basis segment 6 and their corresponding cross-correlation curves and identified peaks.
        -   `get_ccr_peaks(comp, segments, seg_scale, nseg = 1, npeaks = 5)` computes the cross-correlation curve between a basis or scaled segment and the reference signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to `npeaks`, which is an integer. `segments`, `seg_scale`, and `nseg` determines the segment in the cross-correlation computation, and `comp` gives the reference signature. If the multi-segment lengths strategy is used, then `get_ccr_peaks(...)` is called in a `lapply()` for each level of the basis segment. The resulting list is called `ccr.list`.
    +  **multi-segment lengths strategy:** with the multi-segment lengths strategy being used, a position is identified as a candidate position for registration and is called as a "consistent correlation peak" if it results in a top peak in the cross-correlation curve with a tolerance zone determined by `Tx` in all segment levels Note that in @cmps, a segment at its largest scale (highest level) always identifies one peak.
        -   the function `get_seg_scale(segments, nseg, scale = 2)` is used to obtain the scaled version of a basis segment. `segments` is a list containing all basis segments, generated by the function `get_segs(...)` in step 1; `nseg` is an integer indicating which basis segment the function should work with. `scale=1` gives the original length of the basis segment. Increasing `scale` by 1 will double the segment length. 

@@

```{r, eval=FALSE}
extract_feature_cmps(
  x,
  y,
  seg_length = 50,
  Tx = 25,
  npeaks.set = c(5, 3, 1),
  include = NULL
)
```

@@

* `include` determines the format of the function result. Besides the `CMPS.score`, other aspects of the comparison help in understanding how the `CMPS.score` is computed. By default `include` is set to `NULL` and  only  the `CMPS.score` is returned; further results are included when `include` is (an abbreviation of) one of or a vector of the following strings: `"nseg", "congruent.pos", "congruent.seg.idx", "segments", "parameters"`, and `"full_result"`. If `include` is specified as `"full_result"` (or its abbreviation), the ouput includes everything listed below.
  - `nseg`: the number of basis segments  from the comparison signature; this is also the highest possible CMPS score of the comparison;
  - `congruent.pos`: the congruent registration position;
  - `congruent.seg.idx`: the indices of all congruent matching profile segments;
  - `ccp.list`: a list showing identified candidate positions of all basis segments; 
  - `pos.df`: a data frame containing all candidate positions and their respective number of votes;
  - `segments`: a list containing all basis segments;
  - `parameters`: a list containing all input arguments of `extract_feature_cmps`;

@@

comparisons <- comparisons %>% 
  mutate(
    cmps_score = sapply(comparisons$cmps, function(x) x$CMPS.score),
    cmps_nseg = sapply(comparisons$cmps, function(x) x$nseg)
  )

@@

@@

As presented in the work of @cmps, researchers applied the CMPS method to one of the Hamby set scans (Hamby 252). In order to show that our implementation of the CMPS algorithm is able to reproduce the results in @cmps and be used for other data sets, we applied our implementation to both Hamby set 252 and Hamby set 44. Because the CMPS algorithm works with bullet signatures in our case, we conducted some data-processing: for both Hamby 252 and Hamby 44, we started with x3p files in the database, followed the framework proposed by @aoas with the same set of parameters, removed damaged bullet scans, obtained bullet signatures for each bullet land engraving, and removed outliers in bullet signatures. Note that researchers of @cmps used a different framework to obtain bullet signatures. However, since their work is not open-source, we were not able to follow their framework and were only able to reproduce the results for Hamby set 252 qualitatively.

The CMPS score can be considered as a similarity score of two signatures. Then in order to obtain a similarity score of two bullets based on CMPS scores, researchers introduced two similarity metrics, $\mathrm{CMPS_{max}}$ and $\mathrm{\overline{CMPS}_{max}}$, for bullet-level comparisons [@cmps]. $\mathrm{CMPS_{max}}$ is the highest CMPS score obtained among all pairwise bullet signature comparisons. And $\mathrm{\overline{CMPS}_{max}}$ is the highest possible mean CMPS score of signature comparisons that are in the same phase.

In general, we assume each bullet has $n$ land engravings (in our case $n=6$). Let $c_{ij}$ denote the CMPS score of a comparison between bullet 1 land $i$ and bullet 2 land $j$, for $i,j = 1, \dots, n$. Let $\mathcal{P}_k$ denote bullet land pairs in phase $k$ for $k = 0, \dots, n-1$, and $\mathcal{P}_k = \{ (i,j): i = 1, \dots, n ; j = (i + k) - n \cdot \mathbbm{1}_{i + k > n} \}$, where $\mathbbm{1}_A$ denotes an indicator function. For example, $\mathcal{P}_1 = \{ (1,2), (2,3), (3,4), (4,5), (5,6), (6,1) \}$ when $n = 6$.

With that, the two measures to evaluate accuracy used in @cmps are defined as 
\begin{align}
\mathrm{CMPS_{max}} &= \max_{i,j} c_{ij}, \text{and }\\
\mathrm{\overline{CMPS}_{max}} &= \max_{k} \left[  \frac{1}{n} \sum_{(i,j) \in \mathcal{P}_k} c_{ij}\right].
\end{align}
\hh{XXX point out that neither of these measures evaluates the non-matches, and introduce CMPS diff here}

We can continue with the example used in previous sections. `bullets` contains bullet signatures of two bullets, `bullet1` and `bullet2`. As mentioned before, each bullet has six land engravings, resulting in six bullet signatures. Thus, there are 36 pairwise bullet signature comparisons, resulting in 36 $c_{ij}$ values in total. We use multi-segment lengths strategy with default parameters to compute these CMPS scores, and the result is shown in \autoref{fig:tiles}. We can see that in this example,

and since bullet lands in phase $\mathcal{P}_1$ gives the highest mean CMPS score, we have

<!-- based on CMPS scores of bullet signature comparisons. For example, there are in total 6 $\times$ 6 signature comparisons in order to compare two bullets in Hamby study since each bullet has 6 LEAs and each LEA generates a bullet signature. And a CMPS score will be found for every signature comparison. $\mathrm{CMPS_{max}}$ is the highest CMPS score among all signature comparisons. And $\mathrm{\overline{CMPS}_{max}}$ is the highest possible average CMPS score of signature comparisons that are in the same phase. -->

<!-- In this example, $\mathrm{CMPS_{max}}$, the highest CMPS score, is 17, and $\mathrm{\overline{CMPS}_{max}}$, the highest possible mean CMPS score of comparisons in the same phase, is $()/6=12.5$.  -->

>>>>

2. **identify candidate positions:** For each basis segment a set of candidate registration  positions on the comparison signature is identified based on the segment's similarity to the reference signature. In a first step, the cross-correlation function of the segment to the reference is calculated, then a number of positions with high correlation values are identified as candidate positions. In case multiple segment lengths are considered, the length of each basis segment is doubled \wj{by default} and these two steps are repeated. Only when candidate positions coincide (or are similar enough), they are considered further. \autoref{fig:segplots} and \autoref{fig:seg_all} illustrate these ideas.

@@

        -   If `npeaks.set` is an integer vector of length more than 1, for example `npeaks.set = c(5, 3, 1)`, the multi-segment lengths strategy will be used: \wj{calculate the cross-correlation function between a basis segment and the reference and identify positions of the top five peaks; double or increase the legment length to a specified value, recalculate the cross-correlation function, and identify three peaks; repeat this process and identify one peak in the newly computed cross-correlation function.} \autoref{fig:seg_all} shows an example of three levels of basis segment 6 and their corresponding cross-correlation curves and identified peaks. \wj{Note that in @cmps the segment length is doubled at each level of a basis segment, but users are allowed to choose the segment length at each level in our implementation.}
        -   `get_ccr_peaks(comp, segments, seg_outlength, nseg = 1, npeaks = 5)` computes the cross-correlation curve between a basis or increased segment and the reference signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to `npeaks`, which is an integer. `segments`, `seg_outlength`, and `nseg` determines the segment in the cross-correlation computation, and `comp` gives the reference signature. If the multi-segment lengths strategy is used, then `get_ccr_peaks(...)` is called in a `lapply()` for each level of the basis segment. The resulting list is called `ccr.list`.
    +  **multi-segment lengths strategy:** with the multi-segment lengths strategy being used, a position is identified as a candidate position for registration and is called as a "consistent correlation peak" if it results in a top peak in the cross-correlation curve with a tolerance zone determined by `Tx` in all segment levels. Note that in @cmps, a segment at its largest scale (highest level) always identifies one peak \wj{, but we do not have this requirement in our implementation}.
        -   the function `get_seg_scale(segments, nseg, out_length)` is used to obtain the \wj{increased} version of a basis segment. `segments`, which is a list containing all basis segments generated by the function `get_segs(...)` in step 1, and `nseg` are used to determine the basis segment to be increased. `out_length` \wj{specifies the length of the output segment.}

@@

```{r, eval=FALSE}
extract_feature_cmps(
  x,
  y,
  seg_length = 50,
  Tx = 25,
  npeaks.set = c(5, 3, 1),
  include = NULL,
  outlength = NULL
)
```

@@

* `include` determines the format of the function result. Besides the `CMPS.score`, other aspects of the comparison help in understanding how the `CMPS.score` is computed. By default `include` is set to `NULL` and  only  the `CMPS.score` is returned; further results are included when `include` is (an abbreviation of) one of or a vector of the following strings: `"nseg", "congruent.pos", "congruent.seg.idx", "segments", "parameters"`, and `"full_result"`. If `include` is specified as `"full_result"` (or its abbreviation), the output includes everything listed below.
  - `nseg`: the number of basis segments  from the comparison signature; this is also the highest possible CMPS score of the comparison;
  - `congruent.pos`: the congruent registration position;
  - `congruent.seg.idx`: the indices of all congruent matching profile segments;
  - `ccp.list`: a list showing identified candidate positions of all basis segments; 
  - `pos.df`: a data frame containing all candidate positions and their respective number of votes;
  - `segments`: a list containing all basis segments;
  - `parameters`: a list containing all input arguments of `extract_feature_cmps`;

* `outlength` specifies the segment length of a basis segment at each level under the multi-segment lengths strategy. By default `outlength` is set to `NULL`, \wj{indicating that a basis segment should double its segment length at the next level and conforming to the description in} @cmps .

@@

comparisons <- comparisons %>% 
  mutate(
    cmps_score = sapply(comparisons$cmps, function(x) x$CMPS.score),
    cmps_nseg = sapply(comparisons$cmps, function(x) x$nseg),
    cmps_scaled = cmps_score / cmps_nseg
  )

@@

## Methods Maybe?

\wj{The CMPS score can be considered as a similarity score of a land-level comparison, involving two bullet signatures generated from two bullet land engravings. Since each bullet has 6 land engravings in our case, a comparison of two bullets involves 36 land-level comparisons, resulting in 36 CMPS scores (as shown in \autoref{fig:tiles}). In order to obtain a similarity score of a bullet-level comparison summarizing these 36 CMPS scores,} researchers introduced two similarity \wj{metrics (or measures? XXX)}, $\mathrm{CMPS_{max}}$ and $\mathrm{\overline{CMPS}_{max}}$, for bullet-level comparisons [@cmps]. $\mathrm{CMPS_{max}}$ is the highest CMPS score obtained among all land-level comparisons. And $\mathrm{\overline{CMPS}_{max}}$ is the highest possible mean CMPS score of land-level comparisons that are in the same phase.

In general, we assume each bullet has $n$ land engravings (in our case $n=6$). Let $c_{ij}$ denote the CMPS score of a comparison between bullet 1 land $i$ and bullet 2 land $j$, for $i,j = 1, \dots, n$. Let $\mathcal{P}_k$ denote bullet land pairs in phase $k$ for $k = 0, \dots, n-1$, and $\mathcal{P}_k = \{ (i,j): i = 1, \dots, n ; j = (i + k) - n \cdot \mathbbm{1}_{i + k > n} \}$, where $\mathbbm{1}_A$ denotes an indicator function. For example, $\mathcal{P}_1 = \{ (1,2), (2,3), (3,4), (4,5), (5,6), (6,1) \}$ when $n = 6$. Let $k^*$ denote the index of the highest phase.

With that, the two measures to evaluate accuracy used in @cmps are defined as

\begin{align}
\mathrm{CMPS_{max}} &= \max_{i,j} c_{ij} \text{ , and} \\
\mathrm{\overline{CMPS}_{max}} &= \frac{1}{n} \sum_{(i,j) \in \mathcal{P}_{k^*}} c_{ij} \text{ , where} \\
k^* &= \text{arg}\max\limits_{k} \left[  \frac{1}{n} \sum_{(i,j) \in \mathcal{P}_k} c_{ij}\right]
\end{align}

<!-- \begin{align} -->
<!-- \mathrm{CMPS_{max}} &= \max_{i,j} c_{ij}, \text{and }\\ -->
<!-- \mathrm{\overline{CMPS}_{max}} &= \max_{k} \left[  \frac{1}{n} \sum_{(i,j) \in \mathcal{P}_k} c_{ij}\right]. -->
<!-- \end{align} -->
<!-- \hh{XXX point out that neither of these measures evaluates the non-matches, and introduce CMPS diff here} -->

We can continue with the example used in previous sections. `bullets` contains bullet signatures of two bullets, `bullet1` and `bullet2`. As mentioned before, each bullet has 6 land engravings, resulting in 6 bullet signatures. Thus, there are 36 pairwise bullet signature comparisons, resulting in 36 $c_{ij}$ values in total. We use multi-segment lengths strategy with default parameters to compute these CMPS scores, and the result is shown in \autoref{fig:tiles}. We can see that in this example,

and since bullet lands in phase $\mathcal{P}_1$ gives the highest mean CMPS score ($k^* = 1$), we have

However, both $\mathrm{CMPS_{max}}$ and $\mathrm{\overline{CMPS}_{max}}$ consider only relatively high CMPS scores and ignore the rest. So we introduce a new metric based on CMPS scores called $\mathrm{\overline{CMPS}_{diff}}$, which is the difference between $\mathrm{\overline{CMPS}_{max}}$ and the mean of all other CMPS scores. With our notation above, we have:

\begin{align}
\mathrm{\overline{CMPS}_{diff}} = \left[  \frac{1}{n} \sum_{(i,j) \in \mathcal{P}_{k^*}} c_{ij}\right] - \left[  \frac{1}{n(n-1)} \sum_{(i,j) \notin \mathcal{P}_{k^*}} c_{ij}\right]
\end{align}

$\mathrm{\overline{CMPS}_{diff}}$ highlights the difference between CMPS scores of matching and non-matching comparisons. If two bullets are non-matching, all 36 CMPS scores are expected to be small with relatively the same values, resulting in a $\mathrm{\overline{CMPS}_{diff}}$ value close to 0. For the example above, $\mathrm{\overline{CMPS}_{diff}} = 12.5 - 1.53 = 10.97$

**The Scaled CMPS Score**

Another issue with the CMPS score is that the highest possible CMPS score (the total number of basis segments) might differ across comparisons (as shown in \autoref{fig:tiles2}(a)) due to different lengths of bullet signatures and different lengths of basis segments specified by the parameter. A CMPS score of 5 might indicate a non-match if the highest possible CMPS score is 30 but indicate a match if the highest possible CMPS score is 6. Thus, we introduce the scaled CMPS score, denoted as $c^*_{ij}$. Let $s_{ij}$ denote the highest possible CMPS score or the total number of basis segments, we have

\begin{align}
c^*_{ij} = \frac{c_{ij}}{s_{ij}}
\end{align}

The scaled CMPS scores of the above example are shown in \autoref{fig:tiles2}(b). \wj{Compared with the original CMPS scores, the scaled CMPS scores are more comparable in the sense that they all lie within the interval $(0, 1)$ no matter the choice of the length of the basis segments. Note that the scaled CMPS scores can be combined with the CMPS metrics introduced above to obtain their scaled version.} For example,  $\mathrm{\overline{CMPS^*}_{diff}} = 0.498$ for \autoref{fig:tiles2}(b)

```{r tiles2, echo=FALSE, fig.cap="Plot (a) shows the highest possible CMPS scores (the total number of basis segments) for the 36 comparisons. (b) shows the scaled CMPS scores for the 36 comparisons.", out.width=".7\\textwidth", fig.align="center"}
p1 <- dframe %>% ggplot(aes(x = landA, y = landB)) + 
  geom_tile(fill = "grey") + 
  geom_tile(aes(colour="same land"), fill=NA, data = dframe %>% filter(samesource), size=1) + 
  scale_colour_manual("Source", values="darkorange") +
  facet_grid(bulletB ~ bulletA) + xlab("Bullet1 Lands") + 
  ylab("Bullet2 Lands") + 
  geom_text(aes(label=cmps_nseg)) +
  theme_bw() +
  theme(aspect.ratio = 1)

p2 <- dframe %>% ggplot(aes(x = landA, y = landB, fill = cmps_scaled)) + 
  geom_tile() + 
  geom_tile(aes(colour="same land"), fill=NA, data = dframe %>% filter(samesource), size=1) + 
  scale_fill_gradient2("Scaled CMPS", low = "gray80", high = "darkorange", midpoint = 0.3) + 
  scale_colour_manual("Source", values="darkorange") +
  facet_grid(bulletB ~ bulletA) + xlab("Bullet1 Lands") + 
  ylab("Bullet2 Lands") + 
  geom_text(aes(label=round(cmps_scaled, 2))) +
  theme_bw() +
  theme(aspect.ratio = 1)

ggarrange(p1, p2, 
          common.legend = TRUE, legend = "bottom", labels = c("(a)", "(b)"))
```

**Variance Ratio**

We now introduce a measure called "variance ratio" that evaluates the separation between two groups. Let $n_k$ denote the number of observations in group $k$ and $y_{kl}$ denote the $l$-th observation in group $k$, for $k = 1,2$ and $l = 1, \dots, n_k$. Let $\bar{y}_{k.}$ denote the mean value in group $k$ and $\bar{y}_{..}$ denote the overall mean. Then we can define the variance ratio $V$ as:

$$
V = \frac{\sum n_k (\bar{y}_{k.} - \bar{y}_{..})^2}{\sum_k \sum_l^{n_k} (y_{kl} - \bar{y}_{k.})^2 }
$$

This is called "variance ratio" since the numerator evaluates the "variance" between the two groups, and the denominator evaluates the "variance" within each group. \wj{Note that this quantity is not equal to the F-statistic in one-way ANOVA since we ignore the degrees of freedom of both the numerator and the denominator. We do not need the degrees of freedom because all the variance ratio values compared in the next sections have exactly the same number of observations in both goups (the known-match group and the known-nonmatch group), resulting in exactly the same degrees of freedom.}

@@

As presented in the work of @cmps, researchers applied the CMPS method to one of the Hamby set scans (Hamby 252 \wj{to the best of our knowledge} ). In order to show that our implementation of the CMPS algorithm is able to reproduce the results in @cmps and be used for other data sets, we applied our implementation to both Hamby set 252 and Hamby set 44. \wj{For the purpose of reproducibility, we present how we obtained bullet signatures from the Hamby set data}: for both Hamby 252 and Hamby 44, we started with x3p files in the database, followed the framework proposed by @aoas with the same set of parameters, removed damaged bullet scans, obtained bullet signatures for each bullet land engraving, and removed outliers in bullet signatures. \wj{More details and code can be found in the Appendix.} Note that researchers of @cmps \wj{applied the CMPS algorithm to bullet signatures as well but used a framework different from ours.} However, since their work is not open-source, we were not able to follow their framework and were only able to reproduce the results for Hamby set 252 qualitatively.
